---
title: "Getting Started with njcensus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with njcensus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(njcensus)
library(dplyr)
```

## Introduction

The njcensus package provides tools for working with New Jersey Census demographic data. It supports both 2010 and 2020 census years and offers functions for downloading, processing, and analyzing population data by race and gender across counties and municipalities.

## Getting Started

### Initialize the Database

Before retrieving any data, you need to initialize the census database:

``` r
init_census_data()
```

This function:

Creates a DuckDB database file Downloads and processes both 2010 and 2020 census data Sets up the necessary tables for querying

## Basic Data Retrieval

To retrieve data for a specific demographic group and year:

``` r
# Get data for white males in 2020
white_male_2020 <- get_census_data("white", "male", 2020)

# View the first few rows
head(white_male_2020)
```

## Available Demographics

The package supports these demographic groups:

-   white
-   boaa (Black or African American)
-   aian (American Indian and Alaska Native)
-   asian
-   nhpi (Native Hawaiian and Pacific Islander)
-   others
-   two_more (Two or more races)

## Example Analyses

## Population Distribution by County

``` r
# Get Asian population data for 2020
asian_male_2020 <- get_census_data("asian", "male", 2020)
asian_female_2020 <- get_census_data("asian", "female", 2020)

# Combine and analyze by county
asian_by_county <- asian_male_2020 %>%
  full_join(asian_female_2020, by = c("county", "county_name")) %>%
  group_by(county_name) %>%
  summarize(
    total_population = sum(Total_Male, Total_Female, na.rm = TRUE)
  ) %>%
  arrange(desc(total_population))

head(asian_by_county)
```

## Age Distribution Analysis

``` r
# Get data for a specific demographic
data <- get_census_data("white", "male", 2020)

# Analyze age distribution
age_distribution <- data %>%
  select(matches("_years$")) %>%
  summarize(across(everything(), sum)) %>%
  tidyr::pivot_longer(everything(), 
                     names_to = "age_group", 
                     values_to = "count")

# View the first few age groups
head(age_distribution)
```

## Working with Multiple Years

Compare population changes between 2010 and 2020:

``` r
# Get data for both years
data_2010 <- get_census_data("asian", "female", 2010)
data_2020 <- get_census_data("asian", "female", 2020)

# Calculate population change by county
population_change <- data_2010 %>%
  select(county_name, Total_Female) %>%
  rename(pop_2010 = Total_Female) %>%
  full_join(
    data_2020 %>%
      select(county_name, Total_Female) %>%
      rename(pop_2020 = Total_Female),
    by = "county_name"
  ) %>%
  mutate(
    change = pop_2020 - pop_2010,
    percent_change = (change / pop_2010) * 100
  ) %>%
  arrange(desc(percent_change))

head(population_change)
```

## Advanced Usage

### Custom Database Path

You can specify a custom location for the census database:

``` r
# Set custom database path
set_db_path("/path/to/your/census_data.duckdb")

# Get current database path
get_db_path()
```

## Refreshing the Database

To update or rebuild the database:

``` r
init_census_data(force_refresh = TRUE)
```

## Tips and Best Practices

1.  **Memory Management:** The package uses DuckDB for efficient data storage and retrieval. For large queries, consider using dplyr's lazy evaluation.
2.  **Database Connection:** The package handles database connections automatically. You don't need to manage connections manually.
3.  **Error Handling:** All functions include input validation and informative error messages.

## Session Info

``` r
sessionInfo()
```
