---
title: "Working with Population Estimates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Population Estimates}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  eval = FALSE
)
```

```{r setup}
library(njcensus)
library(dplyr)
```

## Introduction to Population Estimates

In addition to decennial census data, the njcensus package now provides access to the Census Bureau's Population Estimates Program data for New Jersey counties. This vignette demonstrates how to access and analyze this data. Accessing Population Estimates Population estimates data can be accessed using the get_pop_estimates() function. By default, this returns age-sex data for all available years and counties.

```{r, eval=FALSE}}
# Get all age-sex estimates
pop_data <- get_pop_estimates()

# View the first few rows
head(pop_data)
```

## Available File Types

The package supports three types of population estimate files:

-   **agesex**: Age and sex data (default)
-   **alldata**: All demographic data including race and ethnicity
-   **syasex**: Single-year age and sex data

You can specify which type you want:

```{r}
# Get all demographic data
demo_data <- get_pop_estimates(file_type = "alldata")

# Get single-year age and sex data
single_year <- get_pop_estimates(file_type = "syasex")
```

## Filtering Data

The function offers several filtering options: 
### By County

```{r}
# Get estimates for Atlantic County
atlantic <- get_pop_estimates(counties = "Atlantic")

# Get estimates for multiple counties
south_jersey <- get_pop_estimates(
  counties = c("Atlantic", "Cape May", "Cumberland")
)
```

### By Year

```{r}
# Get only 2023 estimates
latest <- get_pop_estimates(years = 2023)

# Get estimates for a range of years
recent <- get_pop_estimates(years = c(2022, 2023))
```

## Example Analyses 

## Population Change Over Time

```{r}
# Get all agesex data
pop_data <- get_pop_estimates()

# Calculate county totals by year
county_trends <- pop_data %>%
  filter(AGEGRP == 0) %>%  # Total population (all ages)
  group_by(CTYNAME, YEAR) %>%
  summarize(
    Total = sum(TOT_POP),
    .groups = "drop"
  ) %>%
  arrange(CTYNAME, YEAR)

# View the trends
head(county_trends)
```

## Age Distribution

```{r}
# Get recent population estimates
pop_2023 <- get_pop_estimates(years = 2023)

# Analyze age distribution
age_distribution <- pop_2023 %>%
  filter(AGEGRP > 0) %>%  # Exclude total population row
  group_by(AGEGRP) %>%
  summarize(
    Population = sum(TOT_POP),
    .groups = "drop"
  ) %>%
  mutate(
    Percentage = (Population / sum(Population)) * 100
  )

# View the age distribution
head(age_distribution)
```

## Comparing Census Data with Estimates 

You can also compare decennial census data with more recent population estimates:

```{r}
# Get 2020 census data
census_2020 <- get_census_data("white", "male", 2020) %>%
  group_by(county_name) %>%
  summarize(Census_2020_Total = sum(Total))

# Get 2023 estimates
estimates_2023 <- get_pop_estimates(years = 2023) %>%
  filter(AGEGRP == 0) %>%  # Total population
  group_by(CTYNAME) %>%
  summarize(Estimate_2023_Total = sum(TOT_POP))

# Rename column for joining
estimates_2023 <- estimates_2023 %>%
  rename(county_name = CTYNAME)

# Join and calculate change
comparison <- census_2020 %>%
  left_join(estimates_2023, by = "county_name") %>%
  mutate(
    Change = Estimate_2023_Total - Census_2020_Total,
    Percent_Change = (Change / Census_2020_Total) * 100
  )

# View comparison
head(comparison)
```

## Conclusion 
Population estimates provide valuable data for the years between decennial censuses. The njcensus package makes it easy to access
and analyze this data alongside the more comprehensive decennial census
data.
